---
title: "Data_Preprocessing_For_App"
author: "Vetrie Senthilkumar"
date: "May 20, 2018"
output: html_document
---

Load data and necessary libraries
```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(magrittr)


course_data = read.csv("../UIUC Datasets/gpa/uiuc-gpa-dataset.csv")
gen_eds = read.csv("../UIUC Datasets/geneds/geneds-dataset.csv")
```

Prepare data by calculating Average GPA and Student Enrollment for classes using the grade distribution as well as making some other minor adjustments
```{r}
names(course_data)[7:20] = c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "F", "W")

course_data$Course.Title = as.character(course_data$Course.Title)

course_data$Course_Number = paste(course_data$Subject, course_data$Number)

#Finds total num of students per class by computing sum of grade distributions
course_data$students_enrolled = apply(course_data[, 7:20], MARGIN = 1, sum)

#Calculates average GPA per class using U of I's GPA weighting system
course_data$Average_GPA = (4 * (course_data$`A+`) + 4 * (course_data$A) + 3.67 * (course_data$`A-`) + 3.33 * (course_data$`B+`) + 3 * (course_data$B) + 2.67 * (course_data$`B-`) + 2.33 * (course_data$`C+`) + 2 * (course_data$C) + 1.67 * (course_data$`C-`) + 1.33 * (course_data$`D+`) + 1 * (course_data$D) + .67 * (course_data$`D-`)) / (course_data$students_enrolled - course_data$W)
```

Find overall information for each course using the information given for each of its sections over the years 
```{r}
courses = unique(course_data$Course.Title)

course_info = data.frame(matrix(ncol = 10))
colnames(course_info) = c("Course.Title", "Course.Number", "Student.Enrollment", "A", "B", "C", "D", "F", "W", "Avg.GPA")

for (i in 1 : length(courses)) {
  course_subset = course_data %>% subset(Course.Title == courses[i])
  
  total_enrollment = sum(course_subset$students_enrolled)
  
  A = (sum(course_subset$`A+` + course_subset$A + course_subset$`A-`) / (total_enrollment) * 100)
  B = (sum(course_subset$`B+` + course_subset$B + course_subset$`B-`) / (total_enrollment) * 100) 
  C = (sum(course_subset$`C+` + course_subset$C + course_subset$`C-`) / (total_enrollment) * 100) 
  D = (sum(course_subset$`D+` + course_subset$D + course_subset$`D-`) / (total_enrollment) * 100) 
  Fs = (sum(course_subset$`F`) / (total_enrollment) * 100)
  W = (sum(course_subset$W) / (total_enrollment) * 100)
  
  avg_gpa = sum(course_subset$students_enrolled * course_subset$Average_GPA) / total_enrollment 
  
  info = c(courses[i], course_subset$Course_Number[1], total_enrollment, A, B, C, D, Fs, W, avg_gpa)
  course_info = rbind(course_info, info)
}

course_info[,3:10] = apply(course_info[, 3:10], MARGIN = 2, as.numeric)
course_info$A = paste0(round(course_info$A, 2), "%")
course_info$B = paste0(round(course_info$B, 2), "%")
course_info$C = paste0(round(course_info$C, 2), "%")
course_info$D = paste0(round(course_info$D, 2), "%")
course_info$`F` = paste0(round(course_info$`F`, 2), "%")
course_info$W = paste0(round(course_info$W, 2), "%")
course_info$Avg.GPA = round(course_info$Avg.GPA, 2)
course_info$Subject = str_extract(course_info$Course.Number, pattern = "[[:alpha:]]{1,5}")
course_info = course_info[-1,]
```

Find overall information for each professor by aggregating the information for the different class sections they've taught over the years
```{r}
professors = as.character(unique(course_data$Primary.Instructor))
# professor_names = str_match(professors, pattern = "([[:alpha:]'-~]+).\\s([[:alpha:]]+)(\\s[[:alpha:]])?")

professor_info = data.frame(matrix(ncol = 10))
colnames(professor_info) = c("Instructor", "Course.Number", "Total.Students", "A", "B", "C", "D", "F", "W", "Avg.GPA")

for(i in 1 : length(professors)) {
  professor_subset = course_data %>% subset(Primary.Instructor == professors[i])
  courses_taught = unique(professor_subset$Course_Number)
  
  for (j in 1 : length(courses_taught)) {
    course_subset = professor_subset %>% subset(Course_Number == courses_taught[j])
    A = sum(course_subset$`A+` + course_subset$A + course_subset$`A-`)
    B = sum(course_subset$`B+` + course_subset$B + course_subset$`B-`)
    C = sum(course_subset$`C+` + course_subset$C + course_subset$`C-`)
    D = sum(course_subset$`D+` + course_subset$D + course_subset$`D-`)
    Fs = sum(course_subset$`F`) 
    W = sum(course_subset$W) 
  
    total_enrollment = sum(course_subset$students_enrolled)
    avg_gpa = sum(course_subset$students_enrolled * course_subset$Average_GPA) / total_enrollment 
  
    info = c(professors[i], courses_taught[j], total_enrollment, A, B, C, D, Fs, W, avg_gpa)
    professor_info = rbind(professor_info, info)
  }
}

professor_info[, 3:10] = apply(professor_info[, 3:10], MARGIN = 2, as.numeric)
professor_info$A = paste(round(((professor_info$A / professor_info$Total.Students) * 100), 2), "%")
professor_info$B = paste(round(((professor_info$B / professor_info$Total.Students) * 100), 2), "%")
professor_info$C = paste(round(((professor_info$C / professor_info$Total.Students) * 100), 2), "%")
professor_info$D = paste(round(((professor_info$D / professor_info$Total.Students) * 100), 2), "%")
professor_info$`F` = paste(round(((professor_info$`F` / professor_info$Total.Students) * 100), 2), "%")
professor_info$W = paste(round(((professor_info$W / professor_info$Total.Students) * 100), 2), "%")
professor_info$Avg.GPA = round(professor_info$Avg.GPA, 2)

professor_info = professor_info[-1, ]
```

Add average GPA for each gen ed
```{r}
average_gpas = vector()
enrollments = vector()

for (i in 1 : nrow(gen_eds)) {
  course_subset = course_info %>% subset(Course.Title == gen_eds$Course.Title[i])
  
  if (nrow(course_subset) > 0) {
    average_gpas = c(average_gpas, course_subset$Avg.GPA)
    enrollments = c(enrollments, course_subset$Student.Enrollment)
  } else {
    average_gpas = c(average_gpas, NA)
    enrollments = c(enrollments, NA)
  }

}

gen_eds$Avg.GPA = average_gpas
gen_eds$Student.Enrollment = enrollments
```

Save altered datasets and newly created datasets as .csv files
```{r}
 # write.csv(course_info, "C:/R Projects/UIUC Course Data Analysis/UIUC_Course_Data_App/course_info.csv")

 #  write.csv(professor_info, "C:/R Projects/UIUC Course Data Analysis/UIUC_Course_Data_App/professor_info.csv")

 # write.csv(gen_eds, "C:/R Projects/UIUC Course Data Analysis/UIUC_Course_Data_App/gen_eds_info.csv")
```

Due to some incorrect data entries, different variations of the same instructor's name appear. I attempted to find those duplicate instructors using the code below. After identifying potential duplicate names, I had to manually check if it was a duplicate error and combine the data for the duplicates if it was.  

```{r}
professor_names = sort(unique(professor_info$Instructor))
name_parts = str_match(professor_names, pattern = "([[:alpha:]'-~]+).\\s([[:alpha:]]+)(\\s[[:alpha:]])?")
name_parts[-1, ]
colnames(name_parts) = c("Full", "Last", "First", "Middle")

potential_duplicates = vector()

for(i in 1 : nrow(name_parts) - 1) {
   current_name = name_parts[i, ]
   next_name = name_parts[i + 1, ]
   
   #Consecutive instances of the same last name and similar first names indicate that the first name could be misspelled
   #Consecutive instances of the same first name indicate the last name could be misspelled 
   if (!(NA %in% current_name) & !(NA %in% next_name)) {
      
     if(current_name[2] == next_name[2] & substr(current_name[3],1,3) == substr(next_name[3],1,3)) {
        potential_duplicates = c(potential_duplicates, current_name[1], next_name[1])
      } else if (current_name[3] == next_name[3]){
     potential_duplicates = c(potential_duplicates, current_name[1], next_name[1])
      }
   }
}

potential_duplicates = data.frame(potential_duplicates)
potential_duplicates
```

After viewing the data in the app, I felt that the decimal percentages for the grade distributions were too cluttered and lessened readbility, so I rounded the percents to the nearest integer instead of the nearest hundredth.

```{r}
course_info = read.csv("course_info.csv")
professor_info = read.csv("professor_info.csv")

for (i in 4:9) {
  professor_info[, i] = str_extract(professor_info[, i], pattern = "[0-9]{1,3}[^%]?[0-9]*") %>% as.numeric() %>% round() %>% paste0("%")
  course_info[, i] = str_extract(course_info[, i], pattern = "[0-9]{1,3}[^%]?[0-9]*") %>% as.numeric() %>% round() %>% paste0("%")
}
```

Another feature I decided to add later were course links to U of I's Course Explorer website. Each course has a description on the website that has the following URL format https://courses.illinois.edu/schedule/2018/fall/"Subject"/"Course Level" so I was able to recreate the links using string manipulation

```{r}
course_info = read.csv("course_info.csv")
course_levels = course_info$Course.Number %>% str_extract(pattern = "[0-9]+")

course_info$Links = paste0("https://courses.illinois.edu/schedule/2018/fall/", course_info$Subject, "/", course_levels)
```
